Staged Abstract Interpreters

### 2019-04-04 paper/10_intro.tex
paper revision

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \todo{cite}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. We develop a unified approach to fulfill the four different
semantics, as shown in Figure~\ref{confluence}. Starting with a generic but
parametric interpreter shared by the four semantics, by simply instantiating
the interpreter differently, we can automatically derive a concrete interpreter,
or an abstract interpreter, or their code-generator versions, respectively.

@@

\begin{itemize}
\item Monadic abstractions, which encapsulate the following two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  \todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\end{itemize}

@@

\paragraph{Contributions} The contribution of the paper is twofold:
1) Intellectually, we present a unifying construction that naturally extends the first 
   Futamura projection to abstract interpreters. 
2) Pragmatically, staging abstract interpreters is an approach to improve performance 
   while untouch the soundness. We study the applicability and evaluate the performance 
   of this approach on various analyses.

The paper is organized as follows: \todo{restate this part}

\begin{itemize}
  \item Starting from a generic interface of the interpreter for a functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation (Section~\ref{unstaged_conc}). Then we lift it to the staged
    version by replacing the binding-time types (Section~\ref{stagedinterp}); and to
    the abstract interpreter by applying abstraction to the environment, store, and
    values (Section~\ref{unstaged_abs}). Finally, we show the combination of the two
    transformations, dubbed \textit{staged abstract interpreters}, can be readily
    derived (Section~\ref{sai}).

  \item We demonstrate that if we apply the abstract interpreter on components
    of the program separately, it not only improves the \textit{efficiency} but
    also the \textit{scalability} (Section~\ref{sai} \todo{}), which are two
    major issues in static analysis.

  \item We conduct two case studies to show the applicability and usefulness
    of applying meta-programming to static analysis (Section~\ref{cases_study}).
    (1) We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} technique (Section~\ref{cs_ac}).
    Our approach and AC are able to achieve the same goal, however, as we show,
    one of the advantage of staging through types is it does not require
    whole-analyzer refactoring.
    (2) We extend the staged abstract interpreter to two variants of
    control-flow analysis: context-sensitive pushdown control-flow analysis and
    \todo{store-widened?} (Section~\ref{sai}).

  \item Based on the instantiation of control-flow analysis, we empirically
    evaluate the performance improvement by staging and specialization
    (Section~\ref{evaluation}). We compare the both context-insensitive and
    \note{context-sensitive} analysis.

\end{itemize}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their very recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. We develop a unified approach to fulfill the four different
semantics, as shown in Figure~\ref{confluence}. Starting with a generic but
parametric interpreter shared by the four semantics, by simply instantiating
the interpreter differently, we can automatically derive a concrete interpreter,
or an abstract interpreter, or their code-generator versions, respectively.

@@

\begin{itemize}
\item Monadic abstractions, which encapsulate the following two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\end{itemize}

@@

\paragraph{Contributions} The contribution of the paper is twofold:
1) Intellectually, we present a unifying construction that naturally extends the first 
   Futamura projection to abstract interpreters.
2) Pragmatically, we show that staging abstract interpreters is an approach to
  improve performance while untouch the soundness by case studies and emprical
  evaluation. The paper is organized as follows:

\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{generic_if}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with least engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    \note{garbage collection}, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive, \note{context-sensitive, and store-widening} analysis.
\end{itemize}

### 2019-04-04 paper/20_prelim.tex
misc
paper revision

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one. Then we
will proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}); by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs}).

We choose to use Scala and monad to demonstrate the idea, but the approach
is not restricted to our choice. One can use imperative style and other MSP languages 
(such as MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} 
or Template Haskell \cite{Sheard:2002:TMH:636517.636528} \todo{cite the recent one})
to construct such staged abstract interpreters.

@@

\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \begin{lstlisting}
  type MonadOps[R[_], M[_], A] = {
    def map[B](f: R[A] => R[B]): M[B]
    def flatMap[B](f: R[A] => M[B]): M[B]
  }

  type AnsM[T] <: MonadOps[R, AnsM, T]
  type Ans = AnsM[Value]
    \end{lstlisting}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
// Environment operations
def ask_env: AnsM[Env]
def local_env(ans: Ans)(ρ: R[Env]): Ans
// Store operations
def get_store: AnsM[Store]
def put_store(σ: R[Store]): AnsM[Unit]
def set_store(av: (R[Addr], R[Value])): AnsM[Unit]
    \end{lstlisting}
  \end{subfigure}
\end{figure}
\vspace{-1em}

@@

\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{lstlisting}
            def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
              case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
              case Var(x) => for {                       v <- ev(rhs)
                ρ <- ask_env                             ρ <- ask_env
                σ <- get_store                           α <- alloc(x)
              } yield get(σ, ρ, x)                       _ <- set_store(α → v)
              case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
                ρ <- ask_env                          } yield rt
              } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
              case App(e1, e2) => for {                 v1 <- ev(e1)                                               
                v1 <- ev(e1)                            v2 <- ev(e2)
                v2 <- ev(e2)                          } yield arith(op, v1, v2)
                rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
              } yield rt                                α <- alloc(x)
              case If0(e1, e2, e3) => for {             ρ <- ask_env
                cnd <- ev(e1)                           v <- local_env(ev(rhs))(ρ + (x → α))
                rt <- br0(cnd, ev(e2), ev(e3))          _ <- set_store(α → v)
              } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                      } yield rt                    
            }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}
\vspace{-1em}

>>>>

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monad to demonstrate the
idea, but the approach is not restricted to our choice. One can use
imperative/direct style and other MSP languages (e.g., MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} and Template
Haskell \cite{Sheard:2002:TMH:636517.636528} to construct such staged abstract
interpreters.

\iffalse
Then we
will proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}); by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs}).
\fi

@@

%\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \begin{lstlisting}
  type MonadOps[R[_], M[_], A] = {
    def map[B](f: R[A] => R[B]): M[B]
    def flatMap[B](f: R[A] => M[B]): M[B]
  }

  type AnsM[T] <: MonadOps[R, AnsM, T]
  type Ans = AnsM[Value]
    \end{lstlisting}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.55\textwidth}
    \begin{lstlisting}
// Environment operations
def ask_env: AnsM[Env]
def local_env(ans: Ans)(ρ: R[Env]): Ans
// Store operations
def get_store: AnsM[Store]
def put_store(σ: R[Store]): AnsM[Unit]
def set_store(av: (R[Addr], R[Value])): AnsM[Unit]
    \end{lstlisting}
  \end{subfigure}
\end{figure}
%\vspace{-1em}

@@

%\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{lstlisting}
          def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
            case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
            case Var(x) => for {                      v  <- ev(rhs)
              ρ <- ask_env                            ρ  <- ask_env
              σ <- get_store                          α  <- alloc(x)
            } yield get(σ, ρ, x)                      _  <- set_store(α → v)
            case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
              ρ <- ask_env                          } yield rt
            } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
            case App(e1, e2) => for {                 v1 <- ev(e1)                                               
              v1 <- ev(e1)                            v2 <- ev(e2)
              v2 <- ev(e2)                          } yield arith(op, v1, v2)
              rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
            } yield rt                                α  <- alloc(x)
            case If0(e1, e2, e3) => for {             ρ  <- ask_env
              cnd <- ev(e1)                           v  <- local_env(ev(rhs))(ρ + (x → α))
              rt  <- br0(cnd, ev(e2), ev(e3))         _  <- set_store(α → v)
            } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                    } yield rt                    
          }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}
%\vspace{-1em}

### 2019-04-04 paper/40_absinterp.tex
paper revision

\begin{lstlisting}
  type Result = (R[List[(Value, Store)]], R[Cache])
\end{lstlisting}

This observation leads us to the type definition of @Result@: in concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a list of such pairs
@List[(Value, Store)]@ to represent multiple results. The results is also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

\paragraph{Monads for Abstract Interpretation} Now we incorporate the
nondeterminism and caching into the monad stack. For concrete evaluation, we
already have a reader effect for environments and a state effect for stores. We
further add three more effects based on it: the nondeterminism effect is
represented by the @ListT[M[_], A]@ monad transformer, inside of which, we have
a @ReaderT@ for a @in@ cache, and a @StateT@ for an @out@ cache. \todo{highlight
where changed}

\begin{lstlisting}
  trait AbstractSemantics extends AbstractComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[ListT[ReaderT[StateT[IdM, Cache, ?], Cache, ?], ?], Store, ?], Env, T]
    ...
  }
\end{lstlisting}

The sketch implementation of @ListT@ transformer is shown below. The value @run@
it encapsulates is of type @M[List[A]]@, where @M[_]@ is the inner monad, and
@A@ the element type of the list. The @flatMap@ method first unwraps the inner
monadic value @M[List[A]]@ and obtain a value of @List[A]@, then applies @f@ to
each element in that list, and finally folds the results back to a single
@List[M, B]@.

\begin{lstlisting}
  case class ListT[M[_]: Monad, A](run: M[List[A]]) {
    def flatMap[B](f: A => ListT[M, B]): ListT[M, B] = ...
  }
\end{lstlisting}

\paragraph{Primitive Operations} Some of the primitive operations are changed
according to the new monad stack. A notable change is that the store update
operation joins the value with the existing values, for the same address. Under
abstract interpretation, we may have to explore the results from both branches
of a conditional expression for soundness. In @br0@, we combine the results
using @mplus@ from @MonadPlus@, which requires the values are join-semilattices
(in our case, @List@s and @Map@s).

@@

For function applications, @ap_clo@ looks roughly the same as the concrete
interpreter. The difference is now the function @rator@ is a set of closures, so
we can not directly destruct it by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a list and lifts the elements into the
monad stack. Then we can still implement @ap_clo@ in monadic style, where the
@CloV@ comes from the @ListT@ and the nondeterminism is naturally handled.

\begin{lstlisting}
  def lift_nd[T](vs: List[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    CloV(Lam(x, e), ρ) <- lift_nd[AbsValue](rator.toList)
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

\begin{figure}[t!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for {
            (v, σ) <- lift_nd[(Value, Store)](out(cfg).toList)
             _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e)
            σ <- get_store
            _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\end{figure}

>>>>

\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

This observation leads us to the type definition of @Result@: in concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a set of such pairs
@Set[(Value, Store)]@ to represent multiple results. The results is also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

\paragraph{Monads for Abstract Interpretation} Now we incorporate the
nondeterminism and caching into the monad stack. For concrete evaluation, we
already have a reader effect for environments and a state effect for stores. We
further add three more effects based on it: the nondeterminism effect is
represented by the @SetT[M[_], A]@ monad transformer, inside of which, we have
a @ReaderT@ for a @in@ cache, and a @StateT@ for an @out@ cache. \todo{highlight
where changed}

\begin{lstlisting}
  trait AbstractSemantics extends AbstractComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[SetT[ReaderT[StateT[IdM, Cache, ?], Cache, ?], ?], Store, ?], Env, T]
    ...
  }
\end{lstlisting}

The sketch implementation of @SetT@ transformer is shown below. The value @run@
it encapsulates is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and
@A@ the element type of the list. The @flatMap@ method first unwraps the inner
monadic value @M[SetT[A]]@ and obtain a value of @Set[A]@, then applies @f@ to
each element in that list, and finally folds the results back to a single
@SetT[M, B]@.

\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] = ...
  }
\end{lstlisting}

\paragraph{Primitive Operations} Some of the primitive operations are changed
according to the new monad stack. A notable change is that the store update
operation joins the value with the existing values, for the same address. Under
abstract interpretation, we may have to explore the results from both branches
of a conditional expression for soundness. In @br0@, we combine the results
using @mplus@ from @MonadPlus@, which requires the values are join-semilattices
(in our case, @Set@s and @Map@s).

@@

For function applications, @ap_clo@ looks roughly the same as the concrete
interpreter. The difference is now the function @rator@ is a set of closures, so
we can not directly destruct it by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a set and lifts the elements into the
monad stack. Then we can still implement @ap_clo@ in monadic style, where the
@CloV@ comes from the @SetT@ and the nondeterminism is naturally handled.

\begin{lstlisting}
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    CloV(Lam(x, e), ρ) <- lift_nd[AbsValue](rator)
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

\begin{figure}[t!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for {
            (v, σ) <- lift_nd[(Value, Store)](out(cfg))
             _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e)
            σ <- get_store
            _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\end{figure}

### 2019-04-04 paper/50_stageabs.tex
paper revision

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure as in the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For the readability, we squeeze the three
inner transformers into a single monad @RepListReaderStateM[A, B, C]@, where @A@
and @B@ are both @Cache@ when used in @AnsM@. Now the result type is a pair of
two staged value: @Rep[List[(Value, Store)]]@ and @Rep[Cache]@.

\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type RepListReaderStateM[A, B, C] = RepListT[RepReaderT[RepStateT[RepIdM, A, ?], B, ?], C]
    type AnsM[T] = RepReaderT[RepStateT[RepListReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[List[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

Since our monad stack is uniformly using staged data, so the staged @ListT@ now
stores a staged value of type @Rep[List[A]]@, inside of a staged monad @M@.
\todo{staged ListT implementation}

\begin{lstlisting}
  case class ListT[M[_]: RepMonad, A](run: M[List[A]]) {
    def flatMap[B: Manifest](f: Rep[A] => ListT[M, B]): ListT[M, B] = ...
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(List[(Value,Store)], Cache)]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(List[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f))
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def lift_nd[T](vs: Rep[List[T]]): AnsM[T]
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(List[(Value, Store)], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ <- get_store;     clo <- lift_nd[AbsValue](rator.toList)
    in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(List[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

\vspace{-1em}
\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(List[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg).toList, out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}
\vspace{-1em}

@@

\begin{lstlisting}
  case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(List[(Value, Store)], Cache)]) extends Rep[AbsValue]
  case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                     in: Rep[Cache], out: Rep[Cache]) extends Rep[(List[(Value, Store)], Cache)]
\end{lstlisting}

@@

For example, to fold a singleton list (often appears in @ListT@), e.g.,
@List(x).foldLeft(init)(f)@ where @x@ and @init@ are staged values, a naive code
generator would faithfully apply @foldLeft@ to the list with an anonymous
function. But we can also utilize the algebraic property of @foldLeft@ to
generate cheaper code: @List(x).foldLeft(init)(f) = f(init, x)@. Since the
function @f@ is known at current stage, we completely eliminate the fold
operation and function application. We apply several rewritings enabled by
partial-static data structures, such as @List@ and @Map@, which greatly reduces
the size of residual programs.

>>>>

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure as in the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For the readability, we squeeze the three
inner transformers into a single monad @RepSetReaderStateM[A, B, C]@, where @A@
and @B@ are both @Cache@ when used in @AnsM@. Now the result type is a pair of
two staged value: @Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.

\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type RepSetReaderStateM[A, B, C] = RepSetT[RepReaderT[RepStateT[RepIdM, A, ?], B, ?], C]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

Since our monad stack is uniformly using staged data, so the staged @SetT@ now
stores a staged value of type @Rep[Set[A]]@, inside of a staged monad @M@.
\todo{staged SetT implementation}

\begin{lstlisting}
  case class SetT[M[_]: RepMonad, A](run: M[Set[A]]) {
    def flatMap[B: Manifest](f: Rep[A] => SetT[M, B]): SetT[M, B] = ...
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f))
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def lift_nd[T](vs: Rep[Set[T]]): AnsM[T]
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[(Value, Store)], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ <- get_store;     clo <- lift_nd[AbsValue](rator)
    in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(Set[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

\vspace{-1em}
\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(Set[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg), out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}
\vspace{-1em}

@@

\begin{lstlisting}
  case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value, Store)], Cache)]) extends Rep[AbsValue]
  case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                     in: Rep[Cache], out: Rep[Cache]) extends Rep[(Set[(Value, Store)], Cache)]
\end{lstlisting}

@@

For example, to fold a singleton set (often appears in @SetT@), e.g.,
@Set(x).foldLeft(init)(f)@ where @x@ and @init@ are staged values, a naive code
generator would faithfully apply @foldLeft@ to the set with an anonymous
function. But we can also utilize the algebraic property of @foldLeft@ to
generate cheaper code: @Set(x).foldLeft(init)(f) = f(init, x)@. Since the
function @f@ is known at current stage, we completely eliminate the fold
operation and function application. We apply several rewritings enabled by
partial-static data structures, such as @Set@ and @Map@, which greatly reduces
the size of residual programs.
